Finch: A Datastructure-Driven Array Programming Language

### 2024-03-14 main.tex
some more stuff

- Fortran supported lots of control structures, no data structures, just dense arrays.

- We tried to emulate complex data using only dense arrays but with complicated control structures.

- Recently, people tried to build frameworks to support structured data, but gave up a lot of program side

- We’re bringing these together, both need to work together.

\subsection{Contributions}

\begin{enumerate}
\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
\item <Performance Contributions>
\end{enumerate}

@@

\section{Additional Topics}

\begin{enumerate}
\item Concordant iteration
\item Dimensionalization
\item Bounds analysis/queries
\item Performance warnings

\end{enumerate}

>>>>

%- Fortran supported lots of control structures, no data structures, just dense arrays.
%
%- We tried to emulate complex data using only dense arrays but with complicated control structures.
%
%- Recently, people tried to build frameworks to support structured data, but gave up a lot of program side
%
%- We’re bringing these together, both need to work together.
\subsection{Contributions}

\begin{enumerate}
\item A rich structured array programming language with for-loops
and complex control flow constructs at the same level of productivity
of dense arrays. To our knowledge, the Finch programming language is the first
to support if-conditions, early breaks, and multiple left hand sides over
structured data, as well as complex accesses such as affine indexing or scatter/gather.
\item More complex array structures than ever before. A complete level-by-level
structure-description language for expressing the structure of data
hierarchically. The first such set of formats to efficiently capture banded,
triangular, run-length-encoded, or sparse datasets, and any combination thereof.
\item The Finch compiler specializes programs to data structures in a
predictable, deterministic approach, making it easier to search the complex
space of programs and datastructures to find an appropriate fit for a given
application. The tensor interface makes it easy to extend Finch to new level
formats. A unique tensor lifecycle model enables polymorphism by analyzing the
appropriate stages to insert simple, overloadable, interface functions such as
initialization or finalization. 
\item A high-level array programming language and fusion interface for
operations such as map, broadcast, or reduce that can be compiled to efficient
code using the previous loop-level abstractions.
\item We evaluate the productivity of our language in several case studies,
showing that Finch can be used to accelerate a wide range of applications, 
from classic operations such as spmv and spgemm, to more complex applications such as image processing and graph analytics.
We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
\end{enumerate}

@@

\section{Background}
\subsection{Looplets}
\subsection{FiberTrees}

\subsection{Concordant Iteration}

\subsection{Protocols}

\section{The Finch Language}

\subsection{Syntax and Semantics}

\section{The Finch Compiler}

\subsection{Dimensionalization}

\subsection{Concordization}

\subsection{Bounds Analysis}

\subsection{Performance Warnings}

\subsection{Wrapperization}

\subsection{Simplification and Algebraic Transformations}

\section{The Tensor Interface}

\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}

\subsection{Level Abstraction}

\begin{enumerate}
\item fibers
\item assembly
\item reassembly
\end{enumerate}

\subsection{Core Level Langage Primitives}
\begin{enumerate}
\item SparseList
\item SparseDict
\item ...
\end{enumerate}

\subsection{Wrapper Tensors}

\subsection{Scalars}

\subsubsection{Sparse Scalars}
\subsubsection{Early Break Scalars}

\section{The Finch High-Level API (Needs a Name)}

\subsection{Finch Logic}

\subsection{Finch Interpreter}

\subsection{Lowering}
\subsubsection{Heuristic Optimization}
